---
title: "Installations of packages and software"
output:
  html_document: default
  df_print: paged
  html_notebook: default
---
### Installation of RStudio
Download and install RStudio from the following link (if you already have RStudio, update to the latest version):
<https://www.rstudio.com/products/RStudio/#Desktop>

### Installation of R packages
If asked to update packages (Update all/some/none [a/s/n]) - update all packages (answer a).

-  Bioconductor
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite()
```

The following are Bioconductor packages:

- CAGEr
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("CAGEr")
```

-  DEseq2
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("DESeq2")
```

-  IRanges
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("IRanges")
```

-  GenomicRanges
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("GenomicRanges")
```

-  GenomicFeatures
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("GenomicFeatures")
```

-  AnnotationDbi
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("AnnotationDbi")
```

-  BSgenome
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("BSgenome")
```

-  BSgenome.Drerio.UCSC.danRer7
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("BSgenome.Drerio.UCSC.danRer7")
```

-  ZebrafishDevelopmentalCAGE  
    You will need the Zebrafish CAGE developmental stages, which can be downloaded from:  
    <http://promshift.genereg.net/CAGEr/PackageSource/>

      Download and extract *ZebrafishDevelopmentalCAGE_0.99.0.tar.gz* into your working directory.  
      To install the package use the following code within R:
      
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
install.packages("ZebrafishDevelopmentalCAGE", repos = NULL, type="source")
```

The following are R CRAN packages, also install within R:

-  RColorBrewer
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
install.packages("RColorBrewer", repos="http://cran.rstudio.com/")
```

-  pheatmap
```{r, message=FALSE, warning=FALSE, eval=FALSE, include=TRUE}
install.packages("pheatmap", repos="http://cran.rstudio.com/")
```
  
  
### Software
The following software should be installed from the terminal/command line:
